### 2024-01-05 revision/reply_to_reviewer1.tex

\documentclass[11pt]{article}
\renewcommand*\familydefault{\sfdefault}
%\usepackage{amssymb,amsmath,amsfonts,comment}
%\usepackage{amsmath,amssymb,graphicx,subfigure,psfrag}
\usepackage{amsmath,amssymb,graphicx,subfigure,psfrag,upgreek}
\usepackage{amssymb,mathrsfs}
\usepackage[margin=1in]{geometry}
\usepackage{tikz,pgfplots,graphicx}
\usepackage{color,pdfcolmk}
\usepackage{amsmath}

%\newcommand{\alennote}[1]{\noindent\emph{\textcolor{blue!50!black}{Alen:#1\:}}}
%\newcommand{\gsnote}[1]{\noindent\emph{\textcolor{green}{GS: #1\:}}}
\newcommand{\todo}[1]{\noindent\emph{\textcolor{red}{Todo: #1\:}}}
\newcommand{\note}[1]{\noindent\emph{\vspace{1ex}\textcolor{blue}{Note: #1\:}}\\[1ex]}
\newcommand{\nnote}[1]{\noindent\emph{\vspace{1ex}\textcolor{red}{Note: #1\:}}}
\newcommand{\referee}[1]{\vspace{.1ex}\noindent{\textcolor{blue}{#1}}}
\newcommand{\nnn}{\mathbf{n}}
\newcommand{\fff}{\mathbf{f}}
\newcommand{\uuu}{\mathbf{u}}
\newcommand{\+}{\otimes}
\newcommand{\m}{\mathcal}
\newcommand{\bs}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\eps}{\varepsilon}

\newcommand{\gs}[1]{\textcolor{green}{G: #1}}
\newcommand{\np}[1]{\textcolor{blue}{N: #1}}
\newcommand{\tuck}[1]{\textcolor{brown}{T: #1}}
\newcommand{\mauro}[1]{\textcolor{red}{#1}}

\begin{document}

\section*{Reply to the Reviewer 2:}

Thanks for the careful reading and for your helpful comments and
suggestions.  Please find below point-by-point replies (in black) to
your comments and questions (which are reprinted in blue). To give you
an overview of all the changes in the paper, we also provide a
diff-document that highlights the changes between the initial
submission and this re-submission.\\[1ex]

\referee{This article addresses an important topic in large-to-extreme-scale inverse problems. Namely, efficient approximation of the Hessian of the negative-log of the posterior in Bayesian inversion.
Specifically, this work explores utilizing hierarchical off-diagonal low-rank (HODLR) matrix approximations of the Hessian.
The idea is clearly presented and the manuscript is well-written, however, I believe some---I would say minor---adjustments are required before recommending acceptance for publications.
Please see my comments below. \\}

\paragraph{\sc General Comments:}

\begin{enumerate} 
\item \referee{I do not think the terms ``Gaussianization, Gaussianized, ...'' are properly utilized in this work. The term Gaussianization has been actively used, especially in the machine learning literature, to refer to transformation of non-Gaussian data/variable in to a Gaussian. This requires finding a map/transformation between the original distribution and a Gaussian, e.g., an optimal transport, or normalizing flows. The authors, however, use these terms to denote approximating a non-Gaussian distribution with a localized Gaussian, e.g., following a Laplacian approach.}

\noindent{We appreciate drawing our attention to the fact that our usage of ``Gaussianization, Gaussianized...'', would be unclear to many readers. We have updated our usage to ``the Laplace approximation of the posterior''.}

\item \referee{Re: Algorithm 2 (page 22); it would be great to discuss how $r_1, \ldots, r_{L}$ can be auto tuned. What metric can be used to automatically choose the optimal value of $L$ and then choose the rank for each level automatically.}

\noindent{Done, with regard to choosing the ranks $r_{1},\dots,r_{L}$. We are not aware of an automatic way to choose the optimal value of $L$ but have described the components that we believe could be used to develop such means.}
\end{enumerate}

\paragraph{\sc Other comments:}
\begin{enumerate}
\item  \referee{Page 1, Line 25: ``approximation of Hessian applies...''; Hessian of what?}

\noindent{We now use the more descriptive ``discretized objective functional Hessian vector-products''.}

\item \referee{Page 1, Line 28--31: statement too long, maybe try splitting!}

\noindent{Done.}

\item \referee{Page 1, Line 29: ``sampling from Gaussian proposal distributions'' $\rightarrow$ ``sampling Gaussian distributions''.}

\noindent{We have not used the exact suggested text since the sentence phrasing has been changed. We believe that ``...that provide a means to sample Gaussian distributions...'', is as close to the suggested rephrasing but which maintains consistency with the new sentence structure.}

\item \referee{Page 1, Line 30: I am not sure I understand what the authors mean by ``Newton solves by direct linear methods''!}

\noindent{We had intended to describe methods such as LU factorization, which is known as a direct method of linear system solution. We hope that by referencing a relevant factorization that the message intended to be communicated is now made more clear.}

\item \referee{Page 1, Line 46: ``to inverse problems'' ``with inverse problems''.}

\noindent{Done, thanks!}

\item \referee{Page 2, Line 7: simulation... systems plays'' $\rightarrow$ ``simulations ... systems play''.}

\noindent{Done.}

\item \referee{Page 3, Line 17: ``generate approximate samples from a Gaussian posterior distribution for...'' $\rightarrow$ ``generate samples from an approximate Gaussian posterior for...''}

\noindent{Done, but we have kept ``Gaussian posterior distribution for...''}

\item \referee{Page 3, Line 22: ``including for a Greenland ice...'' $\rightarrow$ ``including a Greenland ice...''}

\noindent{Done.}

\item \referee{Page 3, Line 28: Add a reference; ``In this section,'' $\rightarrow$ ``In this section (2.1),''}

\noindent{Done, but with the capitalization scheme used throughout the manuscript, ``In this Section (2.1),''}

\item \referee{Page 3, line 44: Remove ``i.e., the data,'' as it is redundant.}

\noindent{Done. We have added an additional sentence in which we explicitly state that we will later refer to $\bs{d}$ as the data instead of the more verbose observational data.}

\item \referee{Page 3, Line 46--47: ``and mathematically'' $\rightarrow$ ``and is mathematically''}

\noindent{Done.}

\item \referee{Page 4, Lines 12--13: The authors use the term ``parameter-to-PDE-solution map'' which has not been introduced. Only the combined ``parameter-to-observable map'' has been introduced on line $10$ above. I suggest being consistent and either explicitly decompose the parameter-to-observable-map into a solution operator (solution-to-PDE-solution) and an observation operator, or just restrict usage to ``parameter-to-observable-map''. Either one or both of the solution operator and the observation operator can be nonlinear and each is associated with its own difficulties.}

\noindent{Done. We have restricted usage to the ``parameter-to-observable map''.}

\item \referee{Page 4, Line 27: ``it's gradient'' $\rightarrow$ ``its gradient''}

\noindent{Done.}

\item \referee{Page 4, Lines 28--33: This paragraph requires some
  modification. If you can well-approximate the posterior you do not
  necessarily need MCMC; you can even use naive MC by applying MH step
  to samples from this Gaussian. MCMC does not generally require
  approximating the posterior with a Gaussian, however sampling a
  Gaussian approximation can be useful. For example in Hamiltonian
  Monte Carlo (HMC), samples from a Gaussian approximation can be used
  to build a mass matrix that optimizes the performance of the
  sampler; diagonal of the mass matrix are optimally set to posterior
  precisions (or approximations thereof). See e.g., reference
  \cite{neal2011mcmc, attia2017reduced}.}

\noindent{We agree with the reviewer, but our experience is that using
  MCMC is still superior to using MH (or importance sampling)
  directly, in particular in high dimensions.  We adjusted this
  paragraph in the revised manuscript.}

%\todo{Georg, please see if you you'd like to add or change anything here.}

\item \referee{Page 4, Line 38: define $_{\ell_{2}}$ presented in the previous equation}

\noindent{This item has been addressed by removing any usage of $_{\ell_{2}}$.}

\item \referee{Page 5, Line 38: Maybe redefine paragraph titles to add a period at the end; ``Compression'' $\rightarrow$ ``Compression.''}

\noindent{Done.}

\item \referee{Page 5, Line 50: replace the comma with a semicolon.}

\noindent{Done.}

\item \referee{Page 5, Line 60: ``algorithm is made possible''; where? add a reference/citation here!}

\noindent{Done.}

\item \referee{Page 6, Line 5: a period is missing after ``Approximations''}

\noindent{Done.}

\item \referee{Page 6, Line 6: hereafter use a different symbol for the number of oversampling vectors as $\bs{d}$ has been utilized to denote the data, despite being boldface.}

\noindent{Done. We are now using the symbol $q$ to denote the number of oversampling vectors.}

\item \referee{Page 6, Lines 24--30: this sentence is too long; try splitting.}

\noindent{We have reformulated this sentence.}

\item \referee{Page 7, Line 40: the term data-misfit Hessian has already been introduced before; remove ``i.e., the Hessian ... functional'' as it is redundant.}

\noindent{Done.}

\item \referee{Page 7, Line 44: remove ``of it''}

\noindent{Done.}

\item \referee{Page 8, Line 40: ``basis function'' $\rightarrow$ ``basis functions''}

\noindent{Done.}

\item \referee{Page 8, Line 51: please see my comment above about using the term ``Gaussianized''}

\noindent{Noted and addressed.}

\item \referee{Page 9, Line 22: I am not sure what ``means of applying'' here means!}

\noindent{We have clarified the vague ``means of applying'' with ``means of computing square root and inverse square root matrix-vector products.}

\item \referee{Page 9, Lines 33--40: this paragraph is redundant. This material/argument has already been introduced in Section 2.1.}

\noindent{We believe that this material is important as it helps show how the example problem fits into the larger class of inverse problems, therefore we decided to keep it here.}

%\todo{I (Tucker) believe that this argument could be strengthened.}

\item \referee{Page 9, Line 49: ``can mathematically'' $\rightarrow$ ``can be mathematically''}

\noindent{Done.}

\item \referee{Page 10 I suggest using subequations for the model Equations (6)--(9)}

Done.

\item \referee{Page 10, Lines 16--19, 22, 43: in reference to model equations, use $\backslash$eqref instead of $\backslash$ref}

Done.

\item \referee{Page 11, Line 5: see my comment above about using the term ``Gaussianizaiton''. I will not comment on this hereafter.}

\noindent{Noted and addressed.}

\item \referee{Page 11, Lines 21, 28: use $\backslash$eqref in reference to model equations.}

Done.

\item \referee{Page 11, Line 31: ``add $1\%$ relative Gaussian noise...'': $1\%$ of what? mean true observation magnitude, $\ell_{\infty}$, ...etc.; please clarify!}

\noindent{Done.}

\item \referee{Page 11, Line 57: I do not believe ``approximate samples'' is a correct terminology. I suggest using ``samples from the approximate posterior''.}

\noindent{This has been addressed.}

\item \referee{Page 12, Figure 3 caption: something is not quite right here. The caption states that circles show representative (random measurement locations), however, lines 29-30 on page 11 say that observations are uniformly distributed. What am I missing here? }

\noindent{The figure has been adjusted so that the measurement locations are now illustrated as being uniformly distributed.}

\item \referee{Page 14, Line 57: there is an extra comma after ``more data''}

\noindent{Done.}

\item \referee{Page 15, Figure 7 caption (line 21): ``low-rank (LR)'' $\rightarrow$ ``LR'' as the acronym has already been introduced; same on line 24 of the same caption: ``low-rank'' $\rightarrow$ ``LR'' }

\noindent{Done.}

\item \referee{Page 15, Figure 8 caption: same as the comment above on using ``LR''}

\noindent{Done.}

\item \referee{Page 16, Line 5: remove the comma after ``Albany'', and remove ``so-called''}

\noindent{Done.}

\item \referee{Page 16, Line 30: use $\backslash$eqref in reference to Equation (3)}

Done.

\item \referee{Page 20, Line 29: remove the comma before ``associated to finite...''}

\noindent{Done.}

\item \referee{Page 23, Line  5: ``low-rank'' $\rightarrow$ ``LR'' }

\noindent{Done.}

\end{enumerate}
%\bibliographystyle{unsrt}
\bibliographystyle{iopart-num}
\bibliography{reviewer2_references}

%\noindent [1]- Neal, Radford M. "MCMC using Hamiltonian dynamics." Handbook of markov chain monte carlo 2, no. 11 (2011): 2.

%\noindent [2]- Attia, Ahmed, Razvan Stefanescu, and Adrian Sandu. "The reduced-order hybrid Monte Carlo sampling smoother." International Journal for Numerical Methods in Fluids 83, no. 1 (2017): 28-51.
\end{document}

### 2024-01-05 revision/reply_to_reviewer1.tex

\begin{document}

\section*{Reply to the Reviewer 1:}

Thanks for the careful reading and for your helpful comments and
suggestions.  Please find below point-by-point replies (in black) to
your comments and questions (which are reprinted in blue). To give you
an overview of all the changes in the paper, we also provide a
diff-document that highlights the changes between the initial
submission and this re-submission.\\[1ex]

\referee{Many Hessian-based computational algorithms have been
  developed to solve large-scale inverse problems constrained by
  partial differential equations (PDEs) in the last decade, especially
  in the context of inferring infinite-dimensional
  parameters. However, it remains a critical computational challenge
  to solve such inverse problems when the Hessian of the data-misfit
  term does not have fast spectrum decay, or the Hessian is
  high-rank. This challenge often leads to the need to solve a
  considerable number of PDEs in many (preconditioned) CG iterations
  with an inexact Newton-CG algorithm, one of the most advanced
  optimization algorithms in solving infinite-dimensional inverse
  problems.\\}

  \referee{To address this challenge, this paper proposes an
  efficient preconditioned CG method with the preconditioner computed
  as an approximate Hessian at some suitable parameter point, which is
  demonstrated to significantly reduce the number of required CG
  iterations to achieve given accuracy. The key novelty of this paper
  is the development of an efficient hierarchical matrix approximation
  of the Hessian using only a small number of Hessian matrix-vector
  products, thus a small number of PDE solves. This novelty is made
  possible by exploiting the property of the Hessian operator with
  locally supported non-negative integral kernels. Specifically, the
  authors propose a point spread function (PSF) approximation of the
  Hessian by (1) computing the zeroth, first, and second-order
  impulse response moments of the Hessian by applying its product with
  constant, linear, and quadratic functions, (2) building local
  ellipsoid support es- timate of the impulse response functions of
  the Hessian based on these moments, (3) selecting sample points of
  the impulse response from a candidate set by a greedy ellipsoid
  packing algorithm, (4) computing the impulse responses with disjoint
  ellipsoid in batches by applying the Hessian to Dirac combs, which
  plays a key role in reducing the total number of Hessian
  matrix-vector product, (5) approximating any integral kernel entries
  by a radial basis interpolation based on the computed impulse
  responses, and (6) building the hierarchical matrix approximation of
  the Hessian with the radial basis interpolation of the integral
  kernel entries before applying proper symmetrizing and flipping of
  the negative eigenvalues to make the approximate matrix symmetric
  positive semi-definite.\\}

\referee{The PSF approximation in this paper is built for more general
  operators than Hessian. It demonstrates the effectiveness of the
  PSF-based method by 1 using it to build preconditioners for the
  Hessian operator in two challenging inverse problems of basal
  friction coefficient inversion of ice sheet flow and ini- tial
  condition inversion of advection-dominated transport. It shows
  significant reductions (5-10X) in the required number of PDE solves
  compared to classical regularization-based preconditioning and no
  preconditioning. The paper also presents a comprehensive numerical
  study on the influence of various parameters (data noises, \# batchs,
  diffusion coefficients, terminal times) on the effective- ness and
  data-scalability of the proposed method, showing that the PSF-based
  preconditioners can form good approximations of high-rank Hessians
  using only a small number of operator applications.  The proposed
  method is very interesting and makes a great contribution to solving
  large-scale inverse problems with high-rank Hessians. The
  presentation of the proposed method is concise and illustrative. The
  numerical results are very convincing. Overall, the paper is very
  well written.\\}

\paragraph{General Comments:}

\referee{I recommend to accept the paper for publication with minor
  revisions. I suggest the authors properly address the following
  questions to help for its better understanding and broader
  applications.\\}

\begin{enumerate} 
\item \referee{The author mentioned one limitation of the proposed method:
      the Hessian should have a local non-negative integral kernel,
      which is not satisfied for wave inverse problems that lead to a
      substantial amount of negative entries of the Hessian. It would
      be interesting if the author could elaborate more details with
      intuition or numerical evidence for this property and the
      limitation, especially if you also use Gauss-Newton approximate
      Hessian for the wave inverse problems as in this paper.}

\noindent{To do.}

\item \referee{It is interesting to see the effectiveness of the
  approximate Hessian used as a preconditioner. It would also be very
  interesting to see how accurate the approximate Hessian is compared
  to a full Hessian in terms of the number of batches, ellipsoid
  sizes, and in particular the total number of impulse responses. Can
  you say anything about the convergence of the approximation, either
  numerical or theoretical?}

  \noindent{To do.}

 \item \referee{The paper presented a nice complexity analysis of the steps in
    constructing the approximate Hessian, with the dominating cost
    arising from the Hessian matrix-vector product. It would be
    interesting to see the computational time of each step for the
    numerical examples to provide a good sense of the complexity and
    to support the 5-10X computational reduction by accounting also
    the overhead beyond the PDE solves.}

   \noindent{To do.}

\end{enumerate}

%\bibliographystyle{unsrt}
%\bibliographystyle{iopart-num}
%\bibliography{reviewer2_references}

\end{document}

### 2024-01-05 revision/reply_to_reviewer2.tex

\begin{document}

\section*{Reply to the Reviewer 2:}

Thanks for the careful reading and for your helpful comments and
suggestions.  Please find below point-by-point replies (in black) to
your comments and questions (which are reprinted in blue). To give you
an overview of all the changes in the paper, we also provide a
diff-document that highlights the changes between the initial
submission and this re-submission.\\[1ex]

\referee{I have read the paper with pleasure and have the following
  suggestions for improvement:\\}

\begin{enumerate}
\item \referee{Interesting work, generally well-written. Perhaps a
  small illustrative example with a few figures at the beginning could
  help understand the procedure intuitively.}

@@

  \noindent{To do.}

\item \referee{possibly interesting references for estimating Hessians
  in linearised seismic imaging include:
  \url{https://doi.org/10.1016/j.acha.2007.06.007},
  \url{https://doi.org/10.1190/1.2836323}}

@@

  \noindent{To do.}

\item \referee{How about comparison with other methods, like those
  proposed in [7],[52]? and you say a bit more? it is hard to conclude
  that your method will always outperform these methods.}

\item \referee{Perhaps it is illustrative to include some 1D blurring
  examples to illustrate robustness of method to departure from
  assumptions (positivity, locality) e.g. on 'indefinite' kernels or
  kernels with multiple modes.}

 \item \referee{is code for applying the developed methods on other
   inverse problems available?}

### 2024-01-08 revision/reply_to_reviewer1.tex

\referee{The PSF approximation in this paper is built for more general
  operators than Hessian. It demonstrates the effectiveness of the
  PSF-based method by 1 using it to build preconditioners for the
  Hessian operator in two challenging inverse problems of basal
  friction coefficient inversion of ice sheet flow and ini- tial
  condition inversion of advection-dominated transport. It shows
  significant reductions (5-10X) in the required number of PDE solves
  compared to classical regularization-based preconditioning and no
  preconditioning. The paper also presents a comprehensive numerical
  study on the influence of various parameters (data noises, \# batchs,
  diffusion coefficients, terminal times) on the effective- ness and
  data-scalability of the proposed method, showing that the PSF-based
  preconditioners can form good approximations of high-rank Hessians
  using only a small number of operator applications.  The proposed
  method is very interesting and makes a great contribution to solving
  large-scale inverse problems with high-rank Hessians. The
  presentation of the proposed method is concise and illustrative. The
  numerical results are very convincing. Overall, the paper is very
  well written.\\}

\referee{Thank you for the positive assesment of our work.\\}

@@

\paragraph{General Comments:}

\referee{I recommend to accept the paper for publication with minor
  revisions. I suggest the authors properly address the following
  questions to help for its better understanding and broader
  applications.\\}

\begin{enumerate} 
\item \referee{The author mentioned one limitation of the proposed method:
      the Hessian should have a local non-negative integral kernel,
      which is not satisfied for wave inverse problems that lead to a
      substantial amount of negative entries of the Hessian. It would
      be interesting if the author could elaborate more details with
      intuition or numerical evidence for this property and the
      limitation, especially if you also use Gauss-Newton approximate
      Hessian for the wave inverse problems as in this paper.}

\noindent{Nick will try to get an impulse response for the wave
  problem.}

@@

\item \referee{It is interesting to see the effectiveness of the
  approximate Hessian used as a preconditioner. It would also be very
  interesting to see how accurate the approximate Hessian is compared
  to a full Hessian in terms of the number of batches, ellipsoid
  sizes, and in particular the total number of impulse responses. Can
  you say anything about the convergence of the approximation, either
  numerical or theoretical?}

  \noindent{Nick will try to create a plot that shows some error
    (e.g. Frob norm, cond number) as a function of number of impulse
    responses.}

@@

 \item \referee{The paper presented a nice complexity analysis of the steps in
    constructing the approximate Hessian, with the dominating cost
    arising from the Hessian matrix-vector product. It would be
    interesting to see the computational time of each step for the
    numerical examples to provide a good sense of the complexity and
    to support the 5-10X computational reduction by accounting also
    the overhead beyond the PDE solves.}

   \noindent{Nick will time one of the examples in the
     advection-diffusion.}

### 2024-01-08 revision/reply_to_reviewer2.tex

\item \referee{possibly interesting references for estimating Hessians
  in linearised seismic imaging include:
  \url{https://doi.org/10.1016/j.acha.2007.06.007},
  \url{https://doi.org/10.1190/1.2836323}}

  \noindent{Noemi will give it a try.}

\item \referee{How about comparison with other methods, like those
  proposed in [7],[52]? and you say a bit more? it is hard to conclude
  that your method will always outperform these methods.}

    \noindent{For ref 7: Ask Tucker if we could do HODL on the Hessian
      coming from ice mountain problem and record the number of
      mat-vecs needed to achieve a certain accuracy, something like
      Figure 8 in teh HODLR paper. Nick will do some estimates to
      compare things with ref 52.}

@@

\item \referee{Perhaps it is illustrative to include some 1D blurring
  examples to illustrate robustness of method to departure from
  assumptions (positivity, locality) e.g. on 'indefinite' kernels or
  kernels with multiple modes.}

  \noindent{Maybe we can answer with the wave problem discussion from
    reviewer 1. or even cook up a 2D deblurring problem, such that the
    Hessian entries will be some modified Gaussians.}

### 2024-01-08 revision/reply_to_reviewer2.tex

\item \referee{How about comparison with other methods, like those
  proposed in [7],[52]? and you say a bit more? it is hard to conclude
  that your method will always outperform these methods.}

    \noindent{For ref 7: Ask Tucker if we could do HODL on the Hessian
      coming from ice mountain problem and record the number of
      mat-vecs needed to achieve a certain accuracy, something like
      Figure 8 in teh HODLR paper. Nick will do some estimates to
      compare things with ref 52.}

\item \referee{Perhaps it is illustrative to include some 1D blurring
  examples to illustrate robustness of method to departure from
  assumptions (positivity, locality) e.g. on 'indefinite' kernels or
  kernels with multiple modes.}

  \noindent{Use a 2D-deblurring problem and cook up a Hessian with
    entries given by be some modified Gaussian and show that the error
    increases as more negative entires are present in the kernel.}

@@

 \item \referee{is code for applying the developed methods on other
   inverse problems available?}

  \noindent{Ideally we would like to release the code, however this
    will take time. In the meantime, we will make the code available
    upon request.}

### 2024-01-15 revision/reply_to_reviewer2.tex

\item \referee{possibly interesting references for estimating Hessians
  in linearised seismic imaging include:
  \url{https://doi.org/10.1016/j.acha.2007.06.007},
  \url{https://doi.org/10.1190/1.2836323}}

  \noindent{Noemi will give it a try.}

\item \referee{How about comparison with other methods, like those
  proposed in [7],[52]? and you say a bit more? it is hard to conclude
  that your method will always outperform these methods.}

    \noindent{For ref 7: Ask Tucker if we could do HODLR on the
      Hessian coming from ice mountain problem and record the number
      of mat-vecs needed to achieve a certain accuracy, something like
      Figure 8 in teh HODLR paper. Nick will do some estimates to
      compare things with ref 52.}

### 2024-01-15 paper_sisc_first_revision/localpsf.tex
makefile works
Addressed reviewer 2 point 2

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The method computes impulse responses of the operator at scattered points, and interpolates these impulse responses to approximate entries of the integral kernel. To compute impulse responses efficiently, we apply the operator to Dirac combs associated with batches of point sources, which are chosen by solving an ellipsoid packing problem. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are then performed with fast H-matrix methods. We demonstrate the effectiveness of this PSF-based method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and for the initial condition in an advective-diffusive transport problem. While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. The Hessian impulse responses on the other hand typically become more local as the numerical rank increases, which benefits the PSF-based method. Numerical results reveal that the PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly $5\times$--$10\times$ reductions in the required number of PDE solves, as compared to classical regularization-based preconditioning and no preconditioning. We also present a comprehensive numerical study for the influence of various parameters (that control the shape of the impulse responses and the rank of the Hessian) on the effectiveness of the advection-diffusion Hessian approximation. The results show that the PSF-based preconditioners are able to form good approximations of high-rank Hessians using only a small number of operator applications.
\end{abstract}

@@

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation~\cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems~\cite{AmbartsumyanEtAl20,Hartland_2023}. 
Although these methods are promising,
the required number of operator applications is still large (e.g., hundreds to thousands).
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator applications, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. By using this two stage approach, we reduce the number of operator applications to a few dozen at most. One limitation of our method is that not all data misfit Hessians satisfy the local non-negative integral kernel properties (for example, the wave inverse problem data misfit Hessian has a substantial amount of negative entries in its integral kernel). However, many data misfit Hessians of practical interest do satisfy these properties (either exactly or approximately), and the proposed method is targeted at approximating these Hessians.

>>>>

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The method computes impulse responses of the operator at scattered points, and interpolates these impulse responses to approximate entries of the integral kernel. To compute impulse responses efficiently, we apply the operator to Dirac combs associated with batches of point sources, which are chosen by solving an ellipsoid packing problem. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are then performed with fast H-matrix methods. We demonstrate the effectiveness of this PSF-based method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and for the initial condition in an advective-diffusive transport problem. While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. The Hessian impulse responses on the other hand typically become more local as the numerical rank increases, which benefits the PSF-based method. Numerical results reveal that the PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly $5\times$--$10\times$ reductions in the required number of PDE solves, as compared to classical regularization-based preconditioning and no preconditioning. We also present a comprehensive numerical study for the influence of various parameters (that control the shape of the impulse responses and the rank of the Hessian) on the effectiveness of the advection-diffusion Hessian approximation. The results show that the PSF-based preconditioners are able to form good approximations of high-rank Hessians using only a small number of operator applications.

	TEST ASDF
\end{abstract}

@@

Recently there have been improvements in matrix-free H-matrix construction methods in which an operator is applied to structured random vectors, and the response of the operator to those random vectors is processed to construct an H-matrix approximation~\cite{LevittMartinsson22,LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}. 
These methods (which we do not use here) have been used to approximate Hessians in PDE constrained inverse problems~\cite{AmbartsumyanEtAl20,Hartland_2023}. 
Although these methods are promising,
the required number of operator applications is still large (e.g., hundreds to thousands).
In this paper, we also form an H-matrix Hessian approximation. However, to reduce the required number of operator applications, we first form a PSF approximation of the data misfit Hessian by exploiting locality and non-negative integral kernel properties, then form the H-matrix using classical H-matrix techniques. By using this two stage approach, we reduce the number of operator applications to a few dozen at most.

One limitation of our method is that not all data misfit Hessians satisfy the local non-negative integral kernel properties. We note, in particular, that the wave inverse problem data misfit Hessian has a substantial proportion of negative entries in its integral kernel. In this case more specialized techniques based on pseudodifferential operator theory have been developed, e.g., \cite{DemanetEtAl12,HerrmannMoghaddamStolk08,Symes08}. However, many data misfit Hessians of practical interest do satisfy the local non-negative integral kernel properties (either exactly or approximately), and the proposed method is targeted at approximating these Hessians.
